Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-06 paper/paper2.tex
Finish the last section

%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
\documentclass[journal=jpcbfk, layout=traditional, manuscript=article]{achemso}
\setkeys{acs}{articletitle = true}
\usepackage[utf8]{inputenc}

@@

Two competing sources of error govern the accuracy of an MSM. The first source of error is systematic and is common to all dimensionality reduction methods. \hl{Any simplified representation of a complex system cannot generally describe arbitrary behavior of that system losslessly.} The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories. \hl{As we discuss later,?} These two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}. Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.

@@

The MSM models the \hl{action?} of the propagator with a Markov jump process on $\{1, \ldots, k\}$. Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one. The elements in $\mathbf{p}$ are defined by a coarse-graining of $p(\mathbf{x})$ over $S$, as
\begin{equation}
\label{eq:coarse-grain-p}
p_i = \int_{\mathbf{x} \in s_i} d\mathbf{x}\; p(\mathbf{x}).
\end{equation} 
Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
\begin{equation}
\mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
\end{equation}
where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$. With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$. In a direct analogy with \cref{eq:prop-timescales}, the eigenvalues $\lambda_i$ of $T$ correspond to relaxation timescales of the Markov model
\begin{equation}
\label{eq:msm-timescales}
t_i = -\frac{\tau}{\ln \lambda_i}.
\end{equation}

@@

\section{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}. For example, as we show in appendix A, the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ grows as $O(k)$.

@@

\begin{algorithm}
\caption{Monte Carlo Estimation of the State Volumes}
\begin{algorithmic}[1]
\Procedure{MC Volume Estimation}{$\sigma, A, M$}
	\State $\mathbf{c} \gets [0, 0, \ldots, 0]$\Comment{$\mathbf{c}$ is a vector of length $k$}
	\While{$\sum_{i} c_i \leq M$}
		\State $\mathbf{x} \gets $ sample from axis-aligned hyper cube containing $A$
		\If{$\mathbf{x} \in A$}
			\State $c_{\sigma(\mathbf{x})} \gets c_{\sigma(\mathbf{x})} + 1$
		\EndIf
	\EndWhile
	\State $\mathbf{v} \gets \mathbf{c} \cdot(\sum_{i=1}^k c_i)^{-1}$ \Comment{$\mathbf{v}$ contains the relative volumes of each state}
	\State \textbf{return} $\mathbf{v}$
\EndProcedure
\end{algorithmic}
\end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
\subsection{Simulation Protocol}

@@

How well can likelihood-based methods select parameters for an MSM in the data-rich regime? To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see ``Methods"). We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence. The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states. The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small. For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend. Above $k=600$, the test log-likelihood stops increasing. These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data. On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics. As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=6in]{figs/mull/mull.pdf}
\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

@@

\subsection{Fip35 WW Domain}

Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). \label{fig:ww}}
\end{figure}

For the models built in the tICA subspace, the AIC is comparable to the test set log-likelihood scores using either emission model, displaying a optimum at ~600 states (\cref{fig:ww}). The BIC penalizes complexity more strongly with an optimum at ~150 states. This window is consistent with the convergence of the relaxation timescales and the onset of Markovian behavior. Models built with RMSD require significantly more states than the tICA MSMs and show poor timescale convergence. These results indicate why model selection based on relaxation timescales is so difficult, as a less than optimal state decomposition can result in timescales that do not converge rapidly (if at all). Furthermore, selection based on only the relaxation timescales contains no guard against over-fitting, which is of particular concern in the models built using RMSD that require thousands of states.

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{For MSMs of the WW domain built on the top three PCA components,} 
\end{figure}

%The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models. \citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively. Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered. With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error. On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models. While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande\cite{McGibbon2013Learning}.

\subsection{Limitations and Future Work}

The likelihood functions described herein permit the comparison of Markov state models with varying number of states, however, they require that the compared models have the same support. As such, a direct comparison of the likelihoods between models built with tICA and models built with RMSD is not possible. Furthermore, the uniform emission model is intractable for all but the simplest systems. The Gaussian emission model does not have this limitation.

In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed and the models are parameterized by a direct optimization of \cref{eq:like}. This strategy would complement approaches that generalize MSMs beyond discrete states\cite{Noe2013Variational}.

@@

These results take a step toward fully automating the MSM construction process and controlling the complexity and generalization error by providing a quantitative method for selecting the most suitably complex model, given the data.

@@

\section{Appendix}
\subsection{Other emission distributions}

>>>>

%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
\documentclass[journal=jpcbfk, layout=traditional, manuscript=article]{achemso}
% \setkeys{acs}{articletitle = true}
\usepackage[utf8]{inputenc}

@@

Two competing sources of error govern the accuracy of an MSM. The first source of error is systematic and is common to all dimensionality reduction methods, in that simplified representations of complex systems discard information. The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories. As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}. Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.

@@

The MSM models the dynamics of the propagator with a Markov jump process on $\{1, \ldots, k\}$. Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one. The elements in $\mathbf{p}$ are defined by a coarse-graining of $p(\mathbf{x})$ over $S$, as
\begin{equation}
\label{eq:coarse-grain-p}
p_i = \int_{\mathbf{x} \in s_i} d\mathbf{x}\; p(\mathbf{x}).
\end{equation} 
Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
\begin{equation}
\mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
\end{equation}
where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$. With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ provide approximations for the eigenvalues and eigenfunctions of $\mathcal{Q}$. In a direct analogy with \cref{eq:prop-timescales}, the eigenvalues $\lambda_i$ of $T$ correspond to relaxation timescales of the Markov model
\begin{equation}
\label{eq:msm-timescales}
t_i = -\frac{\tau}{\ln \lambda_i}.
\end{equation}

@@

\section{The Bias-Variance Dilemma}
For MSM methods to be valuable to practitioners, they must operate in limited data regimes. Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters. Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model. This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}. For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.

@@

\begin{algorithm}
\caption{Monte Carlo Estimation of the State Volumes}
\label{alg:volumes}
\begin{algorithmic}[1]
\Procedure{MC Volume Estimation}{$\sigma, A, M$}
	\State $\mathbf{c} \gets [0, 0, \ldots, 0]$\Comment{$\mathbf{c}$ is a vector of length $k$}
	\While{$\sum_{i} c_i \leq M$}
		\State $\mathbf{x} \gets $ sample from axis-aligned hyper cube containing $A$
		\If{$\mathbf{x} \in A$}
			\State $c_{\sigma(\mathbf{x})} \gets c_{\sigma(\mathbf{x})} + 1$
		\EndIf
	\EndWhile
	\State $\mathbf{v} \gets \mathbf{c} \cdot(\sum_{i=1}^k c_i)^{-1}$ \Comment{$\mathbf{v}$ contains the relative volumes of each state}
	\State \textbf{return} $\mathbf{v}$
\EndProcedure
\end{algorithmic}
\end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
\subsection{Simulation Protocol}

@@

How well can likelihood-based methods select parameters for an MSM in the data-rich regime? To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see \emph{Methods}). We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence. The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states. The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small. For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend. Above $k=600$, the test log-likelihood stops increasing. These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data. On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics. As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=6in]{figs/mull/mull.pdf}
\caption{Four model selection criteria were used (left panel) to select the state discretization for MSMs built on simulations of the M\"{u}ller potential. The log-likelihood was evaluated on the training set (black) and a separate test set (blue). These values are plotted as $-2\cdot \log L$ for ease of comparison with the BIC (red) and AIC (purple). Lower scores indicate better models. Dashed lines are used to indicate criterion-minimizing models. These criteria are consistent with the convergence of the implied timescales (right panel).}
\label{fig:mullerlike}
\end{figure}

@@

\subsection{Fip35 WW Domain}
How appropriate are likelihood-based model selection criteria for more realistic MD data sets? To answer this questions, we constructed Markov models based on simulations of the Fip35 WW domain\cite{Liu2008Experimental} performed on the Anton
supercomputer\cite{Shaw2010Atomic}. As a preprocessing step, we used time-structure based independent components analysis (tICA) to project the data into a four dimensional space. This procedure was necessary in order to compute the volumes of the MSM states, as described in \cref{alg:volumes}. MSMs were constructed in this lower dimensional space with differing number of states using the k-centers clustering algorithm (see \emph{Methods}).

As was the case with the M\"{u}ller MSMs, the log-likihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}). The test log likelihood has a maximum at 650 states, and decreases with the addition of further states. This trend is matched by the AIC, which is computed only on the training data, while the BIC on the other hand seems to over-penalize model complexity. The implied timescales of these MSMs converge rapidly, and selection based on their convergence is consistent with the statistical methods.

\begin{figure}
\centering
\label{fig:ww-tica}
\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}

\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria. The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set. Dashed lines are used to indicate criterion-minimizing models. The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly. \hl{right panel should be linear scale instead, for consistency with the PCA}}
\end{figure}

We also built a series of models on the same data set by replacing the tICA preprocessing with Principal Components Analysis (PCA). We projected the data set onto the top three principal components, which are the highest-variance uncorrelated linear combinations of the input degrees of freedom (see \emph{Methods}). Because PCA does not explicitly take into account the temporal structure of the data set, we expected that MSMs based on PCA would be less predictive than those based on tICA. Our results, shown in \cref{fig:ww-pca}, confirm this hypothesis. We built MSMs with as many as 25,000 states, and saw a linear trend in the longest relaxation timescale with respect to the number of states.

The statistical model selection methods indicate that we are only justified in using models with $k$ on the order of 100. The AIC, BIC and test set log-likelihood criteria each recommend a model built with 200 states, which is also the point just after the most dramatic increase in the longest relaxation timescale occurs with respect to $k$. On the other hand, the relaxation timescales continue to increase further, even as the test set log likelihood begins to decrease after 1000 states.

\begin{figure}
\centering
\label{fig:ww-pca}
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA. The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states. The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.} 
\end{figure}

The likelihood functions described herein permit the comparison of Markov state models with varying number of states, however, they require that the compared models have the same support. As such, a direct comparison of the likelihoods between models built with tICA and models built with PCA is not justified.

@@

In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed and the models are parameterized by a direct optimization of \cref{eq:like}. This strategy would complement approaches that generalize MSMs beyond discrete states\cite{chiang2010markov, Noe2013Variational}.

@@

\subsection{Other emission distributions}
